The Sloan Digital Sky Survey Spectroscopic Reduction Pipeline

### 2006-06-09 doc/paper/paper2d.tex
Update.
Fixed some figures

A plate is mounted in the telescope through a efficient procedure described
in Gunn et al. 2006.  The switch from the imaging camera to a spectroscopic
cartridge takes an average of xx minutes, while a switch from one cartridge
to another takes merely x minutes.  Once the cartridge is mounted, the telescope is slewed to the coordinates of the plate center.  
Before acquisition of the guide stars, the 8 flat-field petals are closed and
a sequence of one 10-second flat-field exposure and one 2-second 
Neon-Mercury-Cadmium arc exposure are obtained.  While the arc-lamp frame is
being readout, the flat-field petals are opened, and guide stars are acquired
through the 11 coherent guide bundles (9 are this big and 2 are even bigger??).
With at least 4 guide stars acquired, closed loop guiding is started, and
usually converges to stable guiding within 2 minutes??
The plate scale of the SDSS telescope can be adjusted, and the guiding software
suggest a scale change if one is required.  Once close loop 
guiding has settled, the first science exposure is started.  The typical 
exposure time is 900 seconds, but exposure times from 600 to 1200 second are
not unexpected based on conditions and time constraints.  A second exposure
is started immediately after the first science exposure is complete and readout.A fast linux workstation onsite copies the data from the data acquisition host,
and automatically performs a quick spectral extraction with a subset of the 
steps explained here.  The results are posted to sos.apo.nmsu.edu, and
the SDSS observers can check the S/N obtained in each of the 4 spectroscopic
cameras in real-time.  A plate is considered to have met SDSS survey standards
if all 4 cameras have a cumulative (S/N)$^2$ greater than 15 at 
fidicual magnitudes of ?? and ??.

@@

We utilize an optimal extraction procedure to transform a 2-dimensional image of
the fiber spectrograms to a set of 320 1-dimensional spectra (per spectrograph).
We use the same procedure on all processed frames, 
and will describe the details in processing the flat-field frames, the
arc lamp frames, and finally the science frames.  
The extraction is performed row-by-row, so each frame requires an 
outside loop over 2048 individual rows.
The counts in each row of 2048 columns are modeled with a linear 
combination of 320 profiles plus a low-order
polynomial background.  In addition, deviations to the trace centers and
the profile widths are calculated as linear basis modes (reflecting the first
and second spatial derivatives, respectively).  The compiled routine
we implement to fit the profiles in a single row (extract\_row.c) can
be invoked with any combination of these parameters.  The parameters are
sorted in order of fiber number with background parameters included at the
end of the list.  This facilitates and banded matrix inversion, and 
significantly increases the speed of Cholesky decomposition step in each
iteration of the extraction.

@@

The major obstacle to actual optimal extraction is the determination of 
the exact shape and position of the underlying profile.  
We investigated a range of symmetric, analytic profiles with a 
minimum number of free parameters.  In the end, we found the best 
results using a simple Gaussian or a slightly steeper profile 
characterized by $e^{-|x|^3/\sigma^3}$.  The profiles are defined to be 
normalized to unity, so the amplitude of the profile fit is a 
direct estimator of the counts at a single spectral position.

@@

Each flat-field frame is extracted twice, once assuming a Gaussian
profile and again assuming the steeper $|x|^3$ profile.  In both cases,
the width is allowed to vary freely.  The total number of free parameters
per row in the flat-field extractions is equal to 2 per fiber plus 10
parameters to represent a smooth polynomial background.  After the extractions
are complete, $\chi^2$ is computed by directly comparing the model fit
to the flat-field image.  The median $\chi^2$ for each case is compared
directly, and the better fit and profile (lower median $\chi^2$) are retained
and implemented for the extraction of the remaining frames.  Also retained
is a smooth polynomial fit the average profile width in each fiber bundle
as a function of CCD row number.

@@

\item{Estimate the significance of the residuals in the sky fibers used for the
fit in wavelength bins separated by breakpoints determined above.  For each bin
where inverse the 67th percentile of $\chi^2$ is greater than 1, we rescale 
all the inverse variance values down by the 67th percentile value of $\chi^2$.
The rescaling is actually done with a smooth interpolation as a function of 
redshift with a constraint that the inverse variances cannot increase. }

@@

The accuracy of the SDSS spectral sky subtraction has been tested in recent
scientific analyses using large numbers of background-limited spectra.   
The distribution of flux residuals between spectral observations 
on separate nights has been shown to be 8\% larger in width than the average expected
from the reported errors (McDonald et al. 2004, Burgess \& Burles 2006).
The spectra of faint, background-limited galaxies observed with SDSS exhibit a 
distribution of residuals about the best fit spectral models described by the 
sum of a
Gaussian and Laplace distribution (Bolton et al. 2004).  
The influence of the exponential tail is greatly enhanced in the vicinity of strong sky 
emission lines representative of systematic errors in the brightest lines, but the gaussian 
width closely is closely represented by the reported variance array.

@@

@@

In many of our algorithms we utilize a class of fitting functions known
as B-splines (de Boor 1978).  These are very flexible and efficient 
continous functions which have been applied to numerous spectroscopic
and photometric data sets (cite Kelson ??, Tytler et al. 2004, 
Bolton et al. 2005, Davis/Newman/Finkbeiner) as well as other interpolative
analyses (e.g. Nollett \& Burles 2000).  
We also extend the 1-dimensional
B-splines to slightly higher dimension by allowing the B-spline evalutions
represent polynomial coefficients in a second independent variable.  We 
call these 2-dimensional B-splines, but in fact, 
they are a restricted subset of the full 2-dimensional multi-variate B-splines 
that allow the same degree of freedom in both independent variable domains.

@@

\begin{figure}[tb]
\figurenum{3}
\epsscale{1.0}
\plotone{spDiag2d-0412-52258.ps_1}
\figcaption{A typical example of the common flat-field spectra, $SF$,
for the b1 camera observations of plate 412.  The solid line is the best-fit
B-spline function, and the red points are rejected 2-sigma outliers. 
\label{fig3}
}
\end{figure}

\begin{figure}[tb]
\epsscale{0.9}
\plotone{plot_disp.ps}
\figcaption{Measurements of spectral resolution in the four SDSS 
spectrograph cameras.  Shown for each camera is the median
FWHM arc-line widths as a function of wavelength as 
solid lines for cameras b1 (cyan) and r1 (orange),
dashed lines for cameras b2 (blue) and r2 (red).
The range of resolution is shown as error bars representing
the 10\% to 90\% ranges in each camera over all fibers and
all public data release plates through DR5.
\label{fig_disp}
}
\end{figure}

@@

\begin{figure}
\figurenum{8}
\epsscale{1.0}
\plotone{spDiag2d-0412-52258.ps_156}
\figcaption{An example of 1-d sky subtraction for camera r1 on SDSS
plate 412.  The top panel shows data from 16 sky fibers corrected to airmass
unity with vertical extent equal to 67\% errors.  The solid red line shows
the best fit B-spline, with only wavelength as an independent variable.
The bottom panel shows the residual of the fit in flux counts in the sky 
spectra.
\label{fig8}
}
\end{figure}

\begin{figure}
\figurenum{9}
\epsscale{1.0}
\plotone{spDiag2d-0412-52258.ps_153}
\figcaption{An example of the wavelength corrections to match expected 
sky emission line wavelength centers in a single r1 exposure.  The position
residuals in spectral pixels are shown as measurements of 12 sky 
emission lines for all 320 fibers.  The red positions show the initial 
predicted positions based on the arc-lamp wavelength solutions, and the green
positions are based on the updated solution after shifting and stretching to
match the observed sky lines positions.  The residuals can be compared to a
full pixel in the top left corner, and are much smaller than 5\%.
\label{fig9}
}
\end{figure}

\begin{figure}
\figurenum{10}
\epsscale{1.0}
\plotone{plot_bestthru.ps}
\figcaption{ Throughput of the SDSS spectroscopic system,
as measured relative to the photon density expected above
the Earth's atmospheric by an aperture of 2.5-m with a 1.3-m
obscuration by the secondary mirror baffling.  These curves
are for plate 406 observed on MJD 51817.
Solid curves are for spectrograph \#1, and dotted curves
for spectrograph \#2.
\label{fig10}
}
\end{figure}

\begin{figure}
\figurenum{11}
\epsscale{1.0}
\plottwo{plot_sky_reshist.ps}{plot_lrg_reshist.ps}
\figcaption{ Distribution of SDSS flux residuals normalized 
by the reported pipeline inverse variances.  The left plot shows
the distribution of 214 million sky pixels in reductions from both
Version 4 (upper), and Version 5 (lower) of spectro-2d.  
The best fit functional form is shown as a solid red line (see text).
The filled region shows the area between the observed version 5 distribution 
of sky flux residuals and a gaussian with identical 1-sigma confidence limits.
The right plot is the same but shows the distribution of approximately 
550 million pixels drawn from residuals with respect to best fit models of
luminous red galaxies.
\label{fig11}
}
\end{figure}

>>>>

A plate is mounted in the telescope through a efficient procedure described
in Gunn et al. 2006.  The switch from the imaging camera to a spectroscopic
cartridge takes an average of 30 minutes, while a switch from one cartridge
to another takes 10 minutes.  Once the cartridge is mounted, 
the telescope is slewed to the coordinates of the plate center.  
Before acquisition of the guide stars, the 8 flat-field petals are closed and
a sequence of one 10-second flat-field exposure and one 2-second 
Neon-Mercury-Cadmium arc exposure are obtained.  While the arc-lamp frame is
being readout, the flat-field petals are opened, and guide stars are acquired
through the 11 coherent guide bundles (9 are 7$"$ in diameter and 
2 are even larger at 11$"$).
With at least 4 guide stars acquired, closed loop guiding is started, and
usually converges to stable guiding within 2 minutes.
The plate scale of the SDSS telescope can be adjusted, and the guiding software
suggest a scale change if one is required.  Once close loop 
guiding has settled, the first science exposure is started.  The typical 
exposure time is 900 seconds, but exposure times from 600 to 1200 second are
not unexpected based on conditions and time constraints.  A second exposure
is started immediately after the first science exposure is complete and readout.A fast linux workstation onsite copies the data from the data acquisition host,
and automatically performs a quick spectral extraction with a subset of the 
steps explained here.  The results are posted to sos.apo.nmsu.edu, and
the SDSS observers can check the S/N obtained in each of the 4 spectroscopic
cameras in real-time.  A plate is considered to have met SDSS survey standards
if all 4 cameras have a cumulative (S/N)$^2$ greater than 15 at 
fidicual magnitudes (what the heck is a fiducial magnitude??)
of 20.33 and 20.06 in SDSS g and i filters respectively.

@@

We utilize an optimal extraction procedure to transform a 2-dimensional image of
the fiber spectrograms to a set of 320 1-dimensional spectra (per spectrograph).
We use the same procedure on all processed frames, 
and will describe the details in processing the flat-field frames, the
arc lamp frames, and finally the science frames.  
The extraction is performed row-by-row, so each frame requires an 
outside loop over 2048 individual rows.
The counts in each row of 2048 columns are modeled with a linear 
combination of 320 profiles plus a low-order
polynomial background.  In addition, deviations to the trace centers and
the profile widths are calculated as linear basis modes (reflecting the first
and second spatial derivatives, respectively).  The compiled routine
we implement to fit the profiles in a single row (extract\_row.c) can
be invoked with any combination of these parameters.  The parameters are
sorted in order of fiber number with background parameters included at the
end of the list.  This facilitates a banded matrix inversion, and 
significantly increases the speed of Cholesky decomposition step in each
iteration of the extraction.

@@

The major obstacle to actual optimal extraction is the determination of 
the exact shape and position of the underlying profile.  
We investigated a range of symmetric, analytic profiles with a 
minimum number of free parameters.  In the end, we found the best 
results using a simple Gaussian (noted as PROFTYPE 1 in the primary HDU header) 
or a slightly steeper profile characterized by 
the average of a profile with $e^{-|x|^3/\sigma^3}$ and a Gaussian
with the same sigma (PROFTYPE 3).  The profiles are defined to be 
normalized to unity, so the amplitude of the profile fit is a 
direct estimator of the counts at a single spectral position.

@@

Each flat-field frame is extracted twice, once assuming a Gaussian
profile and again assuming the steeper PROFTYPE 3 profile.  In both cases,
the width is allowed to vary freely.  The total number of free parameters
per row in the flat-field extractions is equal to 2 per fiber plus 10
parameters to represent a smooth polynomial background.  After the extractions
are complete, $\chi^2$ is computed by directly comparing the model fit
to the flat-field image.  The median $\chi^2$ for each case is compared
directly, and the better fit and profile (lower median $\chi^2$) are retained
and implemented for the extraction of the remaining frames.  Also retained
is a smooth polynomial fit the average profile width in each fiber bundle
as a function of CCD row number.

@@

\item{Estimate the significance of the residuals in the sky fibers used for the
fit in wavelength bins separated by breakpoints determined above.  For each bin
where the 67th percentile of $\chi^2$ is greater than 1, we rescale 
all the inverse variance values down by the 67th percentile value of $\chi^2$.
The rescaling is actually done with a smooth interpolation as a function of 
redshift with a constraint that the inverse variances cannot increase. }

@@

The accuracy of the SDSS spectral sky subtraction can been tested by examining
the residuals in final reduced spectra, where spectra of blank sky, high-z QSOs
and luminous red galaxies have been investigated to date.
For early releases of spectroscopic data, reduced by version 4 and include
data releases 4 and earlier, there have been mulitple confirmations of an
underestimation of the true variance as reported in the output of the 
spectroscopic pipeline.  By comparing multiple observations of the same
high-z QSOs in studies of the Lyman-$\alpha$ forest, the average variance was
shown to be underestimated by approximately 16\% (McDonald et al. 2004).  
A fraction of the effect arises due to large tails in the distribution 
of the noise and the remainder from the predicted FWHM of the noise 
distribution.  Switzer (2004) identified two causes for the underestimates,
citing both the awkward treatment of large sky residuals relative to the
supersky and the corresponding correction described above, and also a general
overestimation of the gains applied for the 8 CCD amplifiers.

Using the sample of plates here, we investigate the distribution of the 
noise in a large sample of sky fibers as
well as a sample of model fit luminous red galaxies.  Inclusion of
the latter sample is motivated by earlier studies which showed 
the spectroscopic residuals about the best fit galaxy models 
can be described by a sum of Gaussian and Laplace 
distributions (Bolton et al. 2004).  We construct histograms of the flux
pixels directly from the SDSS spectroscopic products for two versions
of the software (versions 4 \& 5).  In the first sample, flux measurements
from the sky fibers are shown relative to zero and normalized by their reported
inverse variances 
(${\rm residual} = {\rm flux} \, \sqrt{\rm inverse variance}$).  
In the second sample, residuals in galaxy spectra, 
which were targetted as luminous red galaxies, were constructed by subtracting
the best fit galaxy model and then normalizing by the inverse variance.

The results are shown in Figure 11, and also summarized in Table 1.
In Fig. 11, we present the distributions (histograms) as 
probability densities (normalized to unit probability) on a logarithmic 
scale for spectro\_2d versions 4 \& 5.

@@

\bibitem[Schoenberg 1946]{} Schoenberg, I. J. 1946, Quart. Appl. Math., 4, 45

@@

In many of our algorithms we utilize a class of fitting functions known
as B-splines ("basis splines", de Boor 1978, Schoenberg 1946).  
These are very flexible and efficient 
continous functions which have been applied to numerous spectroscopic
and photometric data sets (cite Kelson ??, Tytler et al. 2004, 
Bolton et al. 2005, Davis/Newman/Finkbeiner) as well as other interpolative
analyses (e.g. Nollett \& Burles 2000).  
We also extend the 1-dimensional
B-splines to slightly higher dimension by allowing the B-spline evalutions
represent polynomial coefficients in a second independent variable.  We 
call these 2-dimensional B-splines, but in fact, 
they are a restricted subset of the full 2-dimensional multi-variate B-splines 
that allow the same degree of freedom in both independent variable domains.

@@

\begin{figure}[tb]
\figurenum{3}
\centering
\leavevmode
\includegraphics[width=\columnwidth,clip]{spDiag2d-0412-52258.ps_1}
%\epsscale{1.0}
%\plotone{spDiag2d-0412-52258.ps_1}
\figcaption{A typical example of the common flat-field spectra, $SF$,
for the b1 camera observations of plate 412.  The solid line is the best-fit
B-spline function, and the red points are rejected 2-sigma outliers. 
\label{fig3}
}
\end{figure}

\begin{figure}[tb]
\epsscale{0.9}
\plotone{plot_disp.ps}
\figcaption{Measurements of spectral resolution in the four SDSS 
spectrograph cameras.  Shown for each camera is the median
FWHM arc-line widths as a function of wavelength as 
solid lines for cameras b1 (cyan) and r1 (orange),
dashed lines for cameras b2 (blue) and r2 (red).
The range of resolution is shown as error bars representing
the 10\% to 90\% range in each camera over all fibers and
all public data release plates through DR5.
\label{fig_disp}
}
\end{figure}

@@

\begin{figure}
\figurenum{8}
\epsscale{1.0}
\plotone{skysub_1d.ps}
\figcaption{An example of 1-d sky subtraction for camera r1 on SDSS
plate 412.  The top panel shows data from 16 sky fibers corrected to airmass
unity with vertical extent equal to 67\% errors.  The solid red line shows
the best fit B-spline, with only wavelength as an independent variable.
The bottom panel shows the residual of the fit in flux counts in the sky 
spectra.
\label{fig8}
}
\end{figure}

\begin{figure}
\figurenum{9}
\centering
\leavevmode
\includegraphics[width=\columnwidth,clip]{spDiag2d-0412-52258.ps_153}
%\plotone{spDiag2d-0412-52258.ps_153}
\figcaption{An example of the wavelength corrections to match expected 
sky emission line wavelength centers in a single r1 exposure.  The position
residuals in spectral pixels are shown as measurements of 12 sky 
emission lines for all 320 fibers.  The red positions show the initial 
predicted positions based on the arc-lamp wavelength solutions, and the green
positions are based on the updated solution after shifting and stretching to
match the observed sky lines positions.  The residuals can be compared to a 
full pixel in the top left corner, and are much smaller than 5\% of a pixel
width.
\label{fig9}
}
\end{figure}

\begin{figure}
\figurenum{10}
\epsscale{1.0}
\plotone{plot_bestthru.ps}
\figcaption{ Throughput of the SDSS spectroscopic system,
as measured relative to the photon density expected above
the Earth's atmospheric by an aperture of 2.5-m with a 1.3-m
obscuration by the secondary mirror baffling.  These curves
are for plate 406 observed on MJD 51817, which was observed
in good-seeing photometric conditions shortly after
re-aluminization of the mirrors.
Solid curves are for spectrograph \#1, and dotted curves
for spectrograph \#2.
\label{fig10}
}
\end{figure}

\begin{figure}
\figurenum{11}
\epsscale{1.0}
\plottwo{plot_sky_reshist.ps}{plot_lrg_reshist.ps}
\figcaption{ Distribution of SDSS flux residuals normalized 
by the reported pipeline inverse variances.  The left plot shows
the distribution as a histrogram of 214 million sky pixels in 
reductions from both version 4 (DR4 and earlier, upper) 
and version 5 (lower) of spectro\_2d.  
The best fit functional form is shown as a solid red line (see text).
The filled region shows the area between the observed version 5 distribution 
of sky flux residuals and a gaussian with identical 1-sigma confidence limits.
The right plot is the same but shows the distribution of approximately 
550 million pixels drawn from residuals with respect to best fit models of
luminous red galaxies.
\label{fig11}
}
\end{figure}
